Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-10-17 Manuscript/manuscript.tex
Updates to the results and discussion, not ready

\begin{table}
  \caption{The relative surface excess and the average number of contacts of Calcium with different
    oxygens in POPC with different simulation models \label{tab:binding} }
  \begin{tabular}{l|c c c}
    model                 & $\Gamma_{Ca}^w$ (nm$^{-2}$)   & $\Sigma ^\mathrm{Ca^{2+}} _\mathrm{PO_4} $ & $\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}} $ \\
    \hline
    ECC-lipids            &  $0.07 \pm 0.01 $            &  0.133   &    0.033    \\
    Lipid14/\AA{}qvist    &  $0.13 \pm 0.01 $            &  -       &     -       \\
    Lipid14/Dang          &  $ 0.3 \pm 0.03  $           &  -        &    -       \\
  \end{tabular}
\end{table}
To quantify the binding affinities in different models we calculated 
the relative surface excess from Eq.~\ref{surfexcess}. 
Calcium cations at a molar concentration of $350\,\mathrm{mM}$ provide
$\Gamma_i^w = 0.07 \pm 0.01 \rm{nm}^{-2}$ in the ECC-lipid model, which is significantly
smaller than the values for the Lipid14 model with \AA{}qvist ions, $\Gamma_i^w = 0.13 \pm 0.01 \rm{nm}^{-2}$,
or with Dang ions, $\Gamma_i^w = 0.3 \pm 0.03 \rm{nm}^{-2}$.
\todo{JOE: I'd put the $\Gamma$ values into a table SAMULI: They are now in the table \label{tab:binding} 
with contacts of calcium with different oxygens. However, I am not sure if there is enough information for the
table unless we analyze the contacts for Lipid14 models as well.}
Interestingly, the relative surface excess of 
\ce{NaCl} at $1\,\mathrm{M}$ concentration (ECC-ions~\cite{Pluharova2014}) is qualitatively different from \ce{CaCl2}
having a negative value $\Gamma_i^w = -0.1 \pm 0.01 \rm{nm}^{-2}$, 
meaning that water molecules are preferred to sodium and chloride ions at the membrane-water interface. 
This is in contrast to the most of the available lipid force fields,
which which predict a specific binding of sodium on 
PC lipid bilayers \cite{catte16}.
%\todo{JOE: Add references to NaCl simulation papers that show peaks.
%Add a simple note that there is NBFIX trick in Charmm36 (?).} 
%SAMULI: I think that it is enough to cite NMRlipids II here, which shows all the results
This finding is depicted in Figures~\ref{fig:nacl-dens} and~\ref{fig:cacl-dens},
which show the density profile of ions and water along the membrane normal. 
%The density profiles show
%a larger Ca$^{2+}$ density peak in lipid headgroup region for
%the Lipid14 model with Dang and ECC-ions than for the ECC-lipid model.
\todo{Densities and relative surface excess from NaCl systems to be added.
  The discussion below to be finished after this.}

@@

The contacts between 
%\ce{Ca^{2+}} 
ions and different oxygen moieties of POPC,
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O}$,
were analyzed by calculating the probability of 
having the smaller or equal distance than $0.3 \, \mathrm{nm}$, 
as done previously in Ref.~\citenum{javanainen17}.

@@

and \ce{Na+}
For sodium at $1000\,mM$ concentration 
the average number of contacts per lipid with any oxygen atom in POPC is 
$\Sigma ^\mathrm{Na^{+}} _\mathrm{O} = 0.106 $,    % = 17.1/128 
whereas if only phosphate oxygen atoms are considered 
this number decreases to 
$\Sigma ^\mathrm{Na^{+}} _\mathrm{PO_4} = 0.085 $,    % = 17.0/128 . 
and if only carbonyl oxygen atoms are counted 
it is even lower, 
$\Sigma ^\mathrm{Na^{+}} _\mathrm{O_{carb.}} = 0.048 $.    % = 4.3/128 . 
This means that interaction of sodium and POPC 
has balanced contributions of 55\% from only phosphate oxygens 
(i.e. no contacts with carbonyl oxyens),
and from only carbonyl oxygens with 20\% of configurations 
leaving 25\% for mixed phosphate and carbonyl oxygen configurations.  
This is in line with the calculated negative value of 
relative surface excess of sodium with respect to water, $\Gamma_i^w$, and
their density profiles along the membrane normal (Fig.~\ref{fig:nacl-dens})
suggesting that the populations of bound sodium cations 
naturally decrease with the decreasing density of water.

>>>>

\begin{table}
  \caption{The relative surface excess and the average number of contacts of Calcium with different
    oxygens in POPC with different simulation models \label{tab:binding} }
  \begin{tabular}{l|c c c}
    model                 & $\Gamma_{Ca}^w$ (nm$^{-2}$)   & $\Sigma ^\mathrm{Ca^{2+}} _\mathrm{PO_4} $ & $\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}} $ \\
    \hline
    ECC-lipids            &  $0.07 \pm 0.01 $            &  0.133   &    0.033    \\
    Lipid14/\AA{}qvist    &  $0.13 \pm 0.01 $            &  -       &     -       \\
    Lipid14/Dang          &  $ 0.3 \pm 0.03  $           &  -        &    -       \\
  \end{tabular}
\end{table}
To quantify the binding affinities in different models we calculated 
the relative surface excess from Eq.~\ref{surfexcess}. 
Calcium cations at a molar concentration of $350\,\mathrm{mM}$ provide
$\Gamma_i^w = 0.07 \pm 0.01 \rm{nm}^{-2}$ in the ECC-lipid model, which is significantly
smaller than the values for the Lipid14 model with \AA{}qvist ions, $\Gamma_i^w = 0.13 \pm 0.01 \rm{nm}^{-2}$,
or with Dang ions, $\Gamma_i^w = 0.3 \pm 0.03 \rm{nm}^{-2}$.
\todo{JOE: I'd put the $\Gamma$ values into a table}
Interestingly, the relative surface excess of 
\ce{NaCl} at $1\,\mathrm{M}$ concentration (ECC-ions~\cite{Pluharova2014}) is qualitatively different from \ce{CaCl2}
having a negative value $\Gamma_i^w = -0.1 \pm 0.01 \rm{nm}^{-2}$, 
meaning that water molecules are preferred to sodium and chloride ions at the membrane-water interface. 
This is in contrast to previous works employing classical MD simulations,
which generally agree together on the existence of an increased sodium density at the membrane-water interface.
\todo{JOE: Add references to NaCl simulation papers that show peaks.
Add a simple note that there is NBFIX trick in Charmm36 (?).} 
This finding is depicted in Figures~\ref{fig:nacl-dens} and~\ref{fig:cacl-dens},
which show the density profile of ions and water along the membrane normal. 
%The density profiles show
%a larger Ca$^{2+}$ density peak in lipid headgroup region for
%the Lipid14 model with Dang and ECC-ions than for the ECC-lipid model.
\todo{Densities and relative surface excess from NaCl systems to be added.
  The discussion below to be finished after this.}

@@

We can estimate relative binding affinities 
of several moieties in POPC towards \ce{Ca^{2+}} and \ce{Na+} cations 
using the number of contacts per lipid between cations and POPC oxygen atoms, 
$\Sigma ^\mathrm{Na^{2+}} _\mathrm{O}$,
in which a contact between the two atoms is counted 
if their distance is less than or equal $0.3 \, \mathrm{nm}$ 
(robustness of this definition discussed in reference~\citenum{javanainen17}).

For sodium at $1000\,mM$ concentration 
the average number of contacts per lipid with any oxygen atom in POPC is 
$\Sigma ^\mathrm{Na^{+}} _\mathrm{O} = 0.106 $,    % = 17.1/128 
whereas if only phosphate oxygen atoms are considered 
this number decreases to 
$\Sigma ^\mathrm{Na^{+}} _\mathrm{PO_4} = 0.085 $,    % = 17.0/128 . 
and if only carbonyl oxygen atoms are counted 
it is even lower, 
$\Sigma ^\mathrm{Na^{+}} _\mathrm{O_{carb.}} = 0.048 $.    % = 4.3/128 . 
This means that interaction of sodium and POPC 
has balanced contributions of 55\% from only phosphate oxygens 
(i.e. no contacts with carbonyl oxyens),
and from only carbonyl oxygens with 20\% of configurations 
leaving 25\% for mixed phosphate and carbonyl oxygen configurations.  
This is in line with the calculated negative value of 
relative surface excess of sodium with respect to water, $\Gamma_i^w$, and
their density profiles along the membrane normal (Fig.~\ref{fig:nacl-dens})
suggesting that the populations of bound sodium cations 
naturally decrease with the decreasing density of water.

@@
